Early warning signals: The charted and uncharted territories

### 2013-05-20 manuscript.tex
lots of fun trying to get footnotes without extra latex dependencies

\title{Early warning signals: The charted and uncharted territories}
\author{Carl Boettiger, Noam Ross, Alan Hastings}
\begin{document}

@@

CSD has been studied extensively in theoretical (Wissel 1984, Gandhi et
al. 1998, Carpenter and Brock 2006, Hastings and Wysham 2010, Dakos et
al. 2011a, Lade and Gross 2012, Boettiger and Hastings 2012a) and
empirical contexts (Drake and Griffen 2010, Carpenter 2011a, Veraart et
al. 2012, Dai et al. 2012, Wang et al. 2012) as a potential EWS for
regime shifts. CSD occurs as a system's dominant eigenvalue approaches
zero due to a changing (possibly deteriorating) environment. As the
eigenvalue approaches zero, the system's response to small perturbations
slows. This change in dynamic properties of a system can be expressed in
greater variance, autocorrelation, and return time of observed state
variables.

@@

Another approach to detecting CSD has been fitting time series data to
models. Two approaches have been used for these model-based methods.
First, models may be used to calculate summary statistics related to
CSD, such as eigenvalues (Lade and Gross 2012) or diffusion terms in
jump-diffusion models (Carpenter 2011a, Brock and Carpenter 2012). These
statistics are then examined for trends in the same fashion as the
summary statistics above. Alternatively, models representing both
deteriorating and stable conditions may be fit to the data and in order
to determine which is more likely (Dakos et al. 2012, ). Boettiger and
Hastings (2012a) found that likelihood ratio tests were more powerful
than trend-based summary statistic tests across several real and
simulated ecological data sets. This approach is also more robust than
summary-statistic methods to spurious correlations that arise when
collapses are driven by purely stochastic events (Boettiger and Hastings
2012a).

@@

While saddle-node bifurcations may be common, they represent only part
of the potential mechanisms for rapid regime shift. Occupying the center
of our diagram, Figure 1, such transitions represent our best-understood
cases. Researchers have relied on existing expertise and prior research
to identify empirical systems most likely to experience critical
transitions through the saddle-node-like mechanism (e.g. Carpenter
2011a, Dai et al. 2012), and have achieved a close match to theoretical
predictions of early warning signals. While these examples provide a
much needed proof-of-principle that these signals can be detected in the
real world, it is too early to apply the same methods to novel systems
where the saddle-node is only one of many possible mechanisms. We are
not yet able to determine if a natural system is likely to have a
saddle-node bifurcation without detailed study, despite the popularity
of saddle-node models.

@@

Carpenter, J. 2011a. May the Best Analyst Win. Science 331:698--699.

Carpenter, S. R. 2011a. Early warnings of unknown nonlinear shifts : a
nonparametric approach. Ecology 92:2196--2201.

>>>>

\title{Early warning signals: The charted and uncharted territories}
\author{%
  Carl Boettiger\thanks{CB \& NR contributed equally. Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA} \and Noam Ross\thanks{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA} \and Alan Hastings\footnotemark[2]
}
\begin{document}

\maketitle

@@

CSD has been studied extensively in theoretical (Wissel 1984, Gandhi et
al. 1998, Carpenter and Brock 2006, Hastings and Wysham 2010, Dakos et
al. 2011a, Lade and Gross 2012, Boettiger and Hastings 2012a) and
empirical contexts (Drake and Griffen 2010, Carpenter et al. 2011,
Veraart et al. 2012, Dai et al. 2012, Wang et al. 2012) as a potential
EWS for regime shifts. CSD occurs as a system's dominant eigenvalue
approaches zero due to a changing (possibly deteriorating) environment.
As the eigenvalue approaches zero, the system's response to small
perturbations slows. This change in dynamic properties of a system can
be expressed in greater variance, autocorrelation, and return time of
observed state variables.

@@

Another approach to detecting CSD has been fitting time series data to
models. Two approaches have been used for these model-based methods.
First, models may be used to calculate summary statistics related to
CSD, such as eigenvalues (Lade and Gross 2012) or diffusion terms in
jump-diffusion models (Carpenter 2011, Brock and Carpenter 2012). These
statistics are then examined for trends in the same fashion as the
summary statistics above. Alternatively, models representing both
deteriorating and stable conditions may be fit to the data and in order
to determine which is more likely (Dakos et al. 2012, ). Boettiger and
Hastings (2012a) found that likelihood ratio tests were more powerful
than trend-based summary statistic tests across several real and
simulated ecological data sets. This approach is also more robust than
summary-statistic methods to spurious correlations that arise when
collapses are driven by purely stochastic events (Boettiger and Hastings
2012a).

@@

While saddle-node bifurcations may be common, they represent only part
of the potential mechanisms for rapid regime shift. Occupying the center
of our diagram, Figure 1, such transitions represent our best-understood
cases. Researchers have relied on existing expertise and prior research
to identify empirical systems most likely to experience critical
transitions through the saddle-node-like mechanism (e.g. Carpenter et
al. 2011, Dai et al. 2012), and have achieved a close match to
theoretical predictions of early warning signals. While these examples
provide a much needed proof-of-principle that these signals can be
detected in the real world, it is too early to apply the same methods to
novel systems where the saddle-node is only one of many possible
mechanisms. We are not yet able to determine if a natural system is
likely to have a saddle-node bifurcation without detailed study, despite
the popularity of saddle-node models.

@@

Carpenter, S. R. 2011. Early warnings of unknown nonlinear shifts : a
nonparametric approach. Ecology 92:2196--2201.
